Finch: Sparse and Structured Array Programming with Control Flow

### 2024-04-05 sections/tensors.tex
fixes

\section{Bridging Looplets and Finch: The Tensor Interface}
%
%The Finch language provides descriptions of computations that iterate over a subset of a regular grid that is lexicographically ordered.
%
%Based on our description of Looplets, the reader might believe that compilation of a Finch program simply involves simply replacing for loops over a range with for loops over iterators.
%
%However, Finch's data structures and control flow are sufficiently complex that several challenges exist.
%
Arrays use multiple dimensions to organize data with respect to orthogonal concepts.
%
Thus, the Finch language supports multi-dimensional arrays.
%
Unfortunately, the Looplet abstraction is best suited towards iterators over a single dimension.
%
Our level abstraction provides a bridge between the single dimensional iterators created from looplets and the mutli-dimensional fiber-tree abstractions common to tensor compilers.
% 
This bridge must address three challenges.
%
First, looplets represent an instance of an iterator over a tensor, but we may access the same tensor twice with different indices.
%
Thus, the $unfurl$ function creates separate looplet nests for each iterator.
%
Next, since Finch programs are not just single einsums, they may read and write to the same data at different times.
%
The $declare$, $freeze$, and $thaw$ functions provide machinery to manage transition between these states.
%
Finally, we must be able to write looplet nests that modify tensors, as well as reading them.
%
The $assemble$ function manages the allocation of new data in the tensor.

Additionally, prior fiber-tree representations
focus on sparsity (where only the nonzero elements are represented) and treat
sparse vectors as sets of represented points. Since our fiber-tree
representation must handle other kinds of structure, such as diagonal,
repeated, or constant values, we generalize the concept to view each fiber as a mapping from
indices into a space of subfibers.

In the rest of this section, we discuss $declare$, $freeze$, $thaw$, and $unfurl$ functions as part of a life cycle abstraction as well as the $assemble$ function.
%
These interfaces add to the level abstraction, expanding the types of data that they can express via mapping to Looplets and expanding the contexts in which they can be used.
%
Finally, we describe several levels that we implement in this abstraction, which allow us to implement a taxonomy of known structures that appear in tensor computations.
%
These levels include implementations of various combinations of structures, but they also include specializations to zero dimensional iterators (scalars) and level structures that support different access patterns.

@@

The first three of these functions, declare, freeze, and thaw, \saman{Italics for keywords?}have to do with
managing when tensors can be assumed mutable or immutable.
%
\saman{This is because Finch goes beyond just a single einsum...the scope of Finch w.r.t. TACO etc, is not clear.}
%
As we use Looplets to
represent iteration over a tensor, we must restrict the mutability of tensors
while we iterate over them. 
%
For example, if a tensor declares it has a constant
region from $i = 2:5$, but some other part of the computation modifies the
tensor at $i = 3$, this would result in incorrect behavior.
%
It is much easier to
write correct Looplet code if we can assume that the tensor is immutable while
we are reading from it.
%
Thus, we introduce the notion that a tensor can be in
read-only mode or update-only mode.  
%
In read-only mode, the tensor may only
appear in the right-hand side of assignments
%
. In update-only mode, the tensor
may only appear in the left-hand side of an assignment, either being overwritten
or incremented by some operator. 
%
We can switch between these modes using freeze
and thaw functions.
%
The declare function is used to allocate a tensor,
initialize it to some specified size and value, and leave it in update-only
mode.

>>>>

\section{Bridging Looplets and Finch: The Tensor Interface}
%
%The Finch language provides descriptions of computations that iterate over a subset of a regular grid that is lexicographically ordered.
%
%Based on our description of Looplets, the reader might believe that compilation of a Finch program simply involves simply replacing for loops over a range with for loops over iterators.
%
%However, Finch's data structures and control flow are sufficiently complex that several challenges exist.
%
Arrays use multiple dimensions to organize data with respect to orthogonal concepts.
%
Thus, the Finch language supports multi-dimensional arrays.
%
Unfortunately, the Looplet abstraction is best suited towards iterators over a single dimension.
%
Our level abstraction provides a bridge between the single dimensional iterators created from looplets and the mutli-dimensional fiber-tree abstractions common to tensor compilers.
% 
This bridge must address three challenges.
%
First, while looplets represent an instance of an iterator over a tensor, we may access the same tensor twice with different indices.
%
Thus, the $unfurl$ function creates separate looplet nests for each iterator.
%
Next, since Finch programs go beyond just single einsums, they may read and write to the same data at different times.
%
The $declare$, $freeze$, and $thaw$ functions provide machinery to manage transition between these states.
%
Finally, we must be able to write looplet nests that modify tensors, as well as reading them.
%
The $assemble$ function manages the allocation of new data in the tensor.

Additionally, prior fiber-tree representations focus on sparsity (where only the nonzero elements are represented) and treat sparse vectors as sets of represented points. Since our fiber-tree representation must handle other kinds of structure, such as diagonal, repeated, or constant values, we must generalize our fiber abstraction to allow arbitrary mappings from indices into a space of subfibers.

In the rest of this section, we discuss how these 5 core functions ($declare$, $freeze$, $thaw$, $unfurl$, and $assemble$) function as part of a life cycle abstraction that defines a level in Finch.
%
These interfaces add to the level abstraction, expanding the types of data that they can express via mapping to Looplets and expanding the contexts in which they can be used.
%
We then identify a taxonomy of four key structural properties exhibited in data.
%
We implement several levels in this abstraction that capture all combinations of these structures, 
including specializations to zero dimensional tensors (scalars) and level structures that support different access patterns.

@@

The first three of these functions, $declare$, $freeze$, and $thaw$, have to do with managing when tensors can be assumed mutable or immutable.
%
As we use Looplets to represent iteration over a tensor, we must restrict the mutability of tensors while we iterate over them. 
%
For example, if a tensor declares it has a constant
region from $i = 2:5$, but some other part of the computation modifies the
tensor at $i = 3$, this would result in incorrect behavior.
%
It is much easier to
write correct Looplet code if we can assume that the tensor is immutable while
we are reading from it.
%
Thus, we introduce the notion that a tensor can be in
read-only mode or update-only mode.  
%
In read-only mode, the tensor may only
appear in the right-hand side of assignments
%
. In update-only mode, the tensor
may only appear in the left-hand side of an assignment, either being overwritten
or incremented by some operator. 
%
We can switch between these modes using freeze
and thaw functions.
%
The declare function is used to allocate a tensor,
initialize it to some specified size and value, and leave it in update-only
mode.
